---
title: "DRC2 data cleaning"
author: "Ryan McCord"
date: "2024-09-15"
output:
  html_document:
    df_print: paged
  pdf_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
##Import libraries
library(readxl) #import excel
library(dplyr) #pipes
library(reshape2) #reshaping data
library(lubridate) #to deal with dates
library(tidyverse) #data wrangling
library(ggplot2) #for plotting
library(margins) #for regression analysis
library(vtable) #for dstats
library(plm)
library(stringr)
library(tidyr)
library(ggplot2)
```

## Read in data

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#Create file path
dps2_filepath <- "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DRCDPSdata_Sept2024.xls"

#Import variable names and labels
##var_names <- read_excel(dps2_filepath, range="A1:FU1", col_names = FALSE)
var_labels <- read_excel(dps2_filepath, range="A2:FU2", col_names = FALSE)

#Import data
dps2_RAW <- read_excel(dps2_filepath, skip = 2)

#Assign variable names to data frame
colnames(dps2_RAW) <- as.character(var_labels[1, ]) ##change to var_names when those get added

# #Create data frame with variable names and labels
# var_info <- data.frame(
#   var_names = as.character(var_names[1, ]),
#   var_labels = as.character(var_labels[1, ])
# )

# #Make labels more descriptive
# var_info <- var_info %>%
#   mutate(var_labels = str_replace_all(var_labels, c("A 2.4" = "Mortality - Maternity - ", "D 9.2" = "Diagnosis - IPD - ", "D 9.3" = "Diagnosis - Surgery - ", "D 9.4" = "Diagnosis - Maternity - ", "A 2.5" = "Community - Maternity - ", "A 4.3" = "Equipment functionality - ", "D 4.11" = "Equipment functionality - ", "D 4.12" = "Equipment functionality - ", "F 2.4" = "Equipment functionality - BCZ - ", "D 11.3" = "Procedures - Surgical - ", "D 9.1" = "Diagnosis - Peds - ", "D 9.6" = "Mortality - ")))
```

## Summarize missing values
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Summarize missing values overall and by facility type
missing_summary <- dps2_RAW %>%
  gather(key = "variable", value = "value", -name, -type, -month) %>%
  
  # Calculate the total % missing across all facilities
  group_by(variable) %>%
  summarise(total_missing = round(mean(is.na(value)) * 100, 2)) %>%
  
  # Calculate missing % by facility type and join
  left_join(
    dps2_RAW %>%
      gather(key = "variable", value = "value", -name, -type, -month) %>%
      group_by(variable, type) %>%
      summarise(missing_by_type = round(mean(is.na(value)) * 100, 2), .groups = "drop") %>%
      pivot_wider(names_from = type, values_from = missing_by_type, names_prefix = "missing_by_type_"),
    by = "variable"
  )

# Save the missing summary to a CSV
write.csv(missing_summary, "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DPS2_RAW_missing.csv", row.names = FALSE)
```

## Calcluate avg and max values for each variable at each facility
```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Define the outcome columns (starting from the 4th column)
outcome_columns <- names(dps2_RAW)[4:ncol(dps2_RAW)]

# Calculate facility averages and max values
# Pre-calculate facility averages and max values (renamed to facility_var_stats)
facility_var_stats <- dps2_RAW %>%
  group_by(name) %>%
  mutate(across(all_of(outcome_columns), 
                list(facility_avg = ~ ifelse(all(is.na(.x)), NA, mean(.x, na.rm = TRUE)), 
                     facility_max = ~ ifelse(all(is.na(.x)), NA, max(.x, na.rm = TRUE))))) %>%
  ungroup()

```

## Defining rule to replace zeros and track changes
```{r, echo=FALSE, message=FALSE, warning=FALSE}
zero_replacement <- function(data, outcome_columns, avg_threshold, max_threshold, rule_name) {
  
  for (outcome_col in outcome_columns) {
    
    # Apply the rule using pre-calculated avg and max
    data <- data %>%
      mutate(
        !!paste0("changed_", rule_name, "_", outcome_col) := ifelse(
          is.na(.data[[outcome_col]]) & 
          .data[[paste0(outcome_col, "_facility_avg")]] <= avg_threshold & 
          .data[[paste0(outcome_col, "_facility_max")]] <= max_threshold, 
          TRUE, FALSE),
        
        # Replace missing values if the condition is met
        !!outcome_col := ifelse(
          is.na(.data[[outcome_col]]) & 
          .data[[paste0(outcome_col, "_facility_avg")]] <= avg_threshold & 
          .data[[paste0(outcome_col, "_facility_max")]] <= max_threshold, 
          0, .data[[outcome_col]]
        )
      )
  }
  
  return(data)
}
```

## Apply rule to different combinations of thresholds
```{r, echo=FALSE, message=FALSE, warning=FALSE}
avg_thresholds <- c(2.5, 2.5, 2.5, 3, 3, 3, 3.5, 3.5, 3.5, 4, 4, 4)
max_thresholds <- c(5, 6, 7, 5, 6, 7, 5, 6, 7, 5, 6, 7)

# Initialize a table for results
results <- vector("list", length(avg_thresholds))

# Loop over each threshold combination
for (i in seq_along(avg_thresholds)) {
  avg_thresh <- avg_thresholds[i]
  max_thresh <- max_thresholds[i]
  
  dps2_RAW_test <- zero_replacement(facility_var_stats, outcome_columns, avg_threshold = avg_thresh, max_threshold = max_thresh, rule_name = paste0("avg", avg_thresh, "_max", max_thresh))
  
  # Store the results
  results[[i]] <- data.frame(
    avg_threshold = avg_thresh,
    max_threshold = max_thresh,
    changes_count = sum(dps2_RAW_test[paste0("changed_avg", avg_thresh, "_max", max_thresh, "_", outcome_columns)], na.rm = TRUE)
  )
}

# Combine the results into a single data frame
zero_replacement_summary <- do.call(rbind, results)

# Sort by changes count and write to CSV
zero_replacement_summary <- zero_replacement_summary %>%
  arrange(desc(changes_count))

write.csv(zero_replacement_summary, "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DPS2_RAW_zeros.csv", row.names = FALSE)
```

## Identify variables that change between rule with most (avg 4; max 7) and least changes (avg 2.5; max 5)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
dps2_RAW_with_max7_avg4 <- zero_replacement(facility_var_stats, outcome_columns, avg_threshold = 4, max_threshold = 7, rule_name = "max7_avg4")

dps2_RAW_with_max5_avg2.5 <- zero_replacement(facility_var_stats, outcome_columns, avg_threshold = 2.5, max_threshold = 5, rule_name = "max5_avg2.5")

# Initialize an empty data frame to store the comparison results
variables_changed_df <- data.frame(
  variable = character(),
  changes_count = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each outcome variable to compare changes between the two rules
for (outcome_col in outcome_columns) {
  
  # Get the 'changed' status for each rule
  changed_max7_avg4 <- dps2_RAW_with_max7_avg4[[paste0("changed_max7_avg4_", outcome_col)]]
  changed_max5_avg2.5 <- dps2_RAW_with_max5_avg2.5[[paste0("changed_max5_avg2.5_", outcome_col)]]
  
  # Identify where changes occurred in max7_avg4 but not in max5_avg2.5
  changed_values <- changed_max7_avg4 == TRUE & changed_max5_avg2.5 == FALSE
  
  # Count the number of such changes
  changes_count <- sum(changed_values, na.rm = TRUE)
  
  # If there were changes, add the variable and count to the results dataframe
  if (changes_count > 0) {
    variables_changed_df <- rbind(variables_changed_df, data.frame(
      variable = outcome_col,
      changes_count = changes_count
    ))
  }
}

variables_changed_df <- variables_changed_df %>%
  arrange(desc(changes_count))

# Save the results to a CSV file
write.csv(variables_changed_df, "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DPS2_RAW_zeros_replacementchanges.csv", row.names = FALSE)
```

## Recalculate missing value percentages after applying zero replacement (avg 3; max 6)
```{r, echo=FALSE, message=FALSE, warning=FALSE}
dps2_RAW_int1 <- zero_replacement(facility_var_stats, outcome_columns, avg_threshold = 3, max_threshold = 6, rule_name = "max6_avg3")

# Clean the dataset to remove  '_facility_avg', '_facility_max', and change tracking columns
dps2_RAW_int1_clean <- dps2_RAW_int1 %>%
  select(-ends_with("_facility_avg"), -ends_with("_facility_max"), -starts_with("changed_"))

# Summarize the missing values after zero replacement
missing_summary_updated <- dps2_RAW_int1_clean %>%
  gather(key = "variable", value = "value", -name, -type, -month) %>%
  
  group_by(variable) %>%
  summarise(total_missing = round(mean(is.na(value)) * 100, 2)) %>%
  
  left_join(
    dps2_RAW_int1_clean %>%
      gather(key = "variable", value = "value", -name, -type, -month) %>%
      group_by(variable, type) %>%
      summarise(missing_by_type = round(mean(is.na(value)) * 100, 2), .groups = "drop") %>%
      pivot_wider(names_from = type, values_from = missing_by_type, names_prefix = "missing_by_type_"),
    by = "variable"
  )

write.csv(missing_summary_updated, "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DPS2_missing_zerosreplaced_avg3_max6.csv", row.names = FALSE)

write.csv(dps2_RAW_int1_clean, "C:/Users/rkmcc/Box/2. DRC Energy in Health Facilities/4. Data/DPS2_zerosreplaced_avg3_max6.csv", row.names = FALSE)



```
